---
title: "Clean vaccine dataset"
format:
  html
---

```{r, message=FALSE}
library(here)
library(tidyverse)
library(tinytex)
#library(PupillometryR)
library(DT)
library(lme4)
library(MCMCglmm)
library(brms)
#library(geepack)
library(car)
library(here)
#library(knitr)
#library(kableExtra)
library(table1)
library(readr)
library(dplyr)
library(moments)
#library(sf)
library(tigris)
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
theme_set(theme_bw(base_size = 12)) # Use black/white theme and increase font size for all ggplot figures

```

### Data cleaning

```{r}

## load dataset
vaccine <- read_csv(here("data","vaccine.csv"))


## clean dataset
vaccine_dat <- vaccine %>% 
  
  # recode vaccination coverage rates with -1 as missing
  mutate(overall_recode = ifelse(overall == -1, NA, overall),
         mmr_recode = ifelse(mmr == -1, NA, mmr)) %>%   
  
  # recode missing or 2017 school year as 2017-2018 (as per dataset description)
  mutate(year_recode = case_when(
    year == "null" ~ "2017-18",
    year == "2017" ~ "2017-18",
    TRUE ~ year)) %>%
  
  # recode school type into public and private schools
  mutate(type_recode = case_when(
    type == "BOCES" | type == "Kindergarten" | type == "Public" ~ "Public",
    type == "Charter" | type == "Nonpublic" | type == "Private" ~ "Private",
    TRUE ~ type)) %>%
  
  # create state spending per pupil variable
  mutate(spend_person = round(statespending2016/schagepop2016, 0)) %>%

 # replicate Nhat's data cleaning for county name
  mutate(county = case_when(
    county == "State Of Illinois" ~ NA_character_,
    county == "BLAINE" ~ "Blaine",
    county == "ADA" ~ "Ada",
    county == "#N/A" ~ NA_character_,
    county == "Oklahoma/Tulsa" ~ "Oklahoma",
    county == "Walwroth" ~ "Walworth",
    county == "adams" ~ "Adams",
    county == "Colorado BOCS" ~ "Denver",
    county == "New Hampshire" ~ NA_character_,
    TRUE ~ county)) %>%
  
  # standardize with uppercase
  mutate(county = toupper(county)) %>%
  mutate(state = toupper(state)) %>%

  # create variable for whether the state has county-level data
  mutate(has_county = if_else(!is.na(county), 1, 0)) %>%
    
  # create unique school id
  arrange(state, county, city, name) %>%
  mutate(school_id = row_number())


## create MMR dataset
vaccine_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode)) %>%
  # log10 transformation for negative skew
  mutate(log10_mmr = log10(max(mmr_recode+1) - mmr_recode))

## create overall vaccination dataset
vaccine_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode)) %>%
  # log10 transformation for negative skew
  mutate(log10_overall = log10(max(overall_recode+1) - overall_recode))

saveRDS(vaccine_mmr,file = here("data/vaccine_mmr.rds"))

saveRDS(vaccine_overall,file = here("data/vaccine_overall.rds"))

### for skewness
## resource used for data transformations for skewness: https://www.datanovia.com/en/lessons/transform-data-to-normal-distribution-in-r/
vaccine_overall$overall_recode %>% density() %>% plot()
skewness(vaccine_overall$overall_recode, na.rm = TRUE)

vaccine_overall$log10_overall %>% density() %>% plot()
skewness(vaccine_overall$log10_overall, na.rm = TRUE)

```




$$
E(Y_{ijk}|b^{(3)}_{k},b^{(2)}_{jk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{State electoral college}_k + {b}^{(3)}_k\ + {b}^{(2)}_{jk}\ +\epsilon_{jk}
$$
where Y is vaccination rate (Overall and MMR) for *i*th school in *j*th county and *k*th state (rate),
State spending is in US dollars per student of the *k*th state (continuous),
State electoral college of the *k*th state is a categorical variable (Democrat, Republican, swing state),
${b}_k$ is the random effect of the *k*th state (level 3),
${b}_j$ is the random effect for the *j*th county in the *k*th (level 2).

