---
title: "CHL5222 Group 5 Final Project Report"
subtitle: "R code"
author: "Jannie Leung, Lin Yu, Barrett Buhler, Nhat Chau, Isha Sharma"
format:
  html
---

```{r, message=FALSE}

## load R packages
library(here)
library(tidyverse)
library(tinytex)
library(PupillometryR)
library(DT)
library(lme4)
library(car)
library(knitr)
library(kableExtra)
library(table1)
library(readr)
library(dplyr)
library(sf)
library(tigris)
library(ggpubr)

# set report formatting
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
theme_set(theme_bw(base_size = 12)) # Use black/white theme and increase font size for all ggplot figures

```

In this project, we consulted with the Stackoverflow and Datanovia websites for data analysis.


```{r}

### Data preparation 

## load dataset
vaccine <- read_csv("data/vaccine.csv")

## clean dataset
vaccine_dat <- vaccine %>% 
  
  # recode vaccination coverage rates with -1 as missing
  mutate(overall_recode = ifelse(overall == -1, NA, overall),
         mmr_recode = ifelse(mmr == -1, NA, mmr)) %>%   
  
  # recode missing or 2017 school year as 2017-2018 (as per dataset description)
  mutate(year_recode = case_when(
    year == "null" ~ "2017-18",
    year == "2017" ~ "2017-18",
    TRUE ~ year)) %>%
  
  # create state spending per pupil variable (in US$1million)
  mutate(spend_person = round(statespending2016/schagepop2016/1000, 2)) %>%
  
 # clean county name
  mutate(county = case_when(
    county == "State Of Illinois" ~ NA_character_,
    county == "BLAINE" ~ "Blaine",
    county == "ADA" ~ "Ada",
    county == "#N/A" ~ NA_character_,
    county == "Oklahoma/Tulsa" ~ "Oklahoma",
    county == "Walwroth" ~ "Walworth",
    county == "adams" ~ "Adams",
    county == "Colorado BOCS" ~ "Denver",
    county == "New Hampshire" ~ NA_character_,
    TRUE ~ county)) %>%
  
  # standardize county and state with uppercase
  mutate(county = toupper(county)) %>%
  mutate(state = toupper(state)) %>%
  
  # Add state political electoral college as covariate
  mutate(state_political_vote = case_when(
    state=="ARIZONA" ~ "Swing State",
    state=="ARKANSAS" ~ "Republican Party",
    state=="CALIFORNIA" ~ "Democratic Party",  
    state=="COLORADO" ~ "Democratic Party",     
    state=="CONNECTICUT" ~ "Democratic Party", 
    state=="FLORIDA" ~ "Swing State",      
    state=="IDAHO" ~ "Republican Party",        
    state=="ILLINOIS" ~ "Democratic Party",
    state=="IOWA" ~ "Republican Party",         
    state=="MAINE" ~ "Swing State",  
    state=="MASSACHUSETTS" ~ "Democratic Party", 
    state=="MICHIGAN" ~ "Republican Party", 
    state=="MINNESOTA" ~ "Swing State", 
    state=="MISSOURI" ~ "Republican Party", 
    state=="MONTANA" ~ "Republican Party",     
    state=="NEW JERSEY" ~ "Democratic Party",  
    state=="NEW YORK" ~ "Democratic Party", 
    state=="NORTH CAROLINA" ~ "Swing State", 
    state=="NORTH DAKOTA" ~ "Republican Party", 
    state=="OHIO" ~ "Republican Party",         
    state=="OKLAHOMA" ~ "Republican Party",     
    state=="OREGON" ~ "Democratic Party",       
    state=="PENNSYLVANIA" ~ "Swing State", 
    state=="RHODE ISLAND" ~ "Democratic Party", 
    state=="SOUTH DAKOTA" ~ "Republican Party", 
    state=="TENNESSEE" ~ "Republican Party",    
    state=="TEXAS" ~ "Republican Party", 
    state=="UTAH" ~ "Republican Party", 
    state=="VERMONT" ~ "Democratic Party",      
    state=="VIRGINIA" ~ "Democratic Party",     
    state=="WASHINGTON" ~ "Democratic Party",  
    state=="WISCONSIN" ~ "Swing State"
    )) %>%
  
    # assign Swing State as reference category
    mutate(state_political_vote = factor(state_political_vote, levels = c(
      "Swing State",
      "Democratic Party",
      "Republican Party"))) %>%
    
  # create unique school id
  arrange(state, county, city, name) %>%
  mutate(school_id = row_number())


## create dataset for schools with overall vaccination rates
vaccine_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode))

## create dataset for schools with MMR vaccination rates
vaccine_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode))

```

### Descriptive statistics

```{r}
total_data <- vaccine_dat %>%
  filter(mmr_recode != -1 | overall_recode != -1) %>%  
  summarise(
    private_total = sum(type_recode == "Private", na.rm = TRUE),
    public_total = sum(type_recode == "Public", na.rm = TRUE),
    na_total = sum(is.na(type_recode) | type_recode == " ", na.rm = TRUE))
```

```{r}
### Table 1 summary

state_count_mmr <- vaccine_dat %>% 
  filter(!is.na(mmr_recode)) %>% 
  summarise(n_states = n_distinct(state))
state_count_overall <- vaccine_dat %>% 
  filter(!is.na(overall_recode)) %>% 
  summarise(n_states = n_distinct(state))

county_stats_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode)) %>%
  group_by(state) %>%
  summarise(n_counties = n_distinct(county)) %>%
  summarise(
    total = sum(n_counties),
    median = median(n_counties),
    min = min(n_counties),
    max = max(n_counties)
  )

county_stats_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode)) %>%
  group_by(state) %>%
  summarise(n_counties = n_distinct(county)) %>%
  summarise(
    total = sum(n_counties),
    median = median(n_counties),
    min = min(n_counties),
    max = max(n_counties)
  )

school_stats_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode)) %>%
  group_by(state) %>%
  group_by(year_recode) %>% 
  summarise(n_schools = n_distinct(name)) %>%
  summarise(
    total = sum(n_schools),
    median = median(n_schools),
    min = min(n_schools),
    max = max(n_schools)
  )

school_stats_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode)) %>%
  group_by(state) %>%
  summarise(n_schools = n_distinct(name)) %>%
  summarise(
    total = sum(n_schools),
    median = median(n_schools),
    min = min(n_schools),
    max = max(n_schools)
  )


spending_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode)) %>% 
  group_by(state) %>% 
  summarise(spend = unique(spend_person)) %>% 
  ungroup() %>% 
  summarise(mean_spend_thousand = round(mean(spend, na.rm = TRUE), 2),
            sd_spend_thousand = round(sd(spend, na.rm = TRUE), 2))


spending_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode)) %>% 
  group_by(state) %>% 
  summarise(spend = unique(spend_person)) %>% 
  ungroup() %>% 
  summarise(mean_spend_thousand = round(mean(spend, na.rm = TRUE), 2),
            sd_spend_thousand = round(sd(spend, na.rm = TRUE), 2))

mean_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode)) %>%
  summarise(mean_mmr = round(mean(mmr_recode, na.rm = TRUE), 1),
            sd_mmr = round(sd(mmr_recode, na.rm = TRUE), 1))

mean_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode)) %>%
  summarise(mean_overall = round(mean(overall_recode, na.rm = TRUE), 2),
            sd_overall = round(sd(overall_recode, na.rm = TRUE), 2))

library(tibble)

summary_table <- tibble(
  `Outcome Variable` = c("MMR", "Overall"),
  `State (Count)` = c(
    n_distinct(vaccine_dat$state[!is.na(vaccine_dat$mmr_recode)]),
    n_distinct(vaccine_dat$state[!is.na(vaccine_dat$overall_recode)])
),

`County Count (Sum)` = c(
  paste(county_stats_mmr$total),
  paste(county_stats_overall$total)
),

`Median County per State (Min, Max)` = c(
  paste(county_stats_mmr$median, 
        "[", county_stats_mmr$min, ",", county_stats_mmr$max, "]"),
  paste(county_stats_overall$median, 
        "[", county_stats_overall$min, ",", county_stats_overall$max, "]")
),

`School Count (Sum)` = c(
  paste(school_stats_mmr$total),
  paste(school_stats_overall$total)
),

 `Median School per State (Min, Max)` = c(
  paste(school_stats_mmr$median, 
        "[", school_stats_mmr$min, ",", school_stats_mmr$max, "]"),
  paste(school_stats_overall$median, 
        "[", school_stats_overall$min, ",", school_stats_overall$max, "]")
),

  `Average Spending per Person in $1000 USD` = c(
    paste(spending_mmr$mean_spend_thousand, "(", spending_mmr$sd_spend_thousand, ")"),
    paste(spending_overall$mean_spend_thousand, "(", spending_overall$sd_spend_thousand, ")")
),

  `Schools in 2017–18` = c(
    table(vaccine_dat$year_recode[!is.na(vaccine_dat$mmr_recode)])["2017-18"],
    table(vaccine_dat$year_recode[!is.na(vaccine_dat$overall_recode)])["2017-18"]
),

  `Schools in 2018–19` = c(
    table(vaccine_dat$year_recode[!is.na(vaccine_dat$mmr_recode)])["2018-19"],
    table(vaccine_dat$year_recode[!is.na(vaccine_dat$overall_recode)])["2018-19"]
),

  `Mean Vaccination Rate (%)` = c(
    paste(mean_mmr$mean_mmr, "(", mean_mmr$sd_mmr, ")"),
    paste(mean_overall$mean_overall, "(", mean_overall$sd_overall, ")")
))

summary_table %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

### Scatter plots

We explored the distribution of the averaged vaccination rates.

```{r}
avg_overall_dat <- full_join(
  vaccine_overall %>% 
  group_by(state) %>% 
  summarise(avg_overall = mean(overall)),

vaccine_overall %>% 
  select(state,spend_person,statespending2016) %>% unique(),
by= c("state")
)

avg_overall_dat %>% datatable()

avg_mmr_dat <- full_join(
  vaccine_mmr %>% 
  group_by(state) %>% 
  summarise(avg_mmr = mean(mmr)),

vaccine_mmr %>% 
  select(state,spend_person,statespending2016) %>% unique(),
by= c("state")
)
avg_mmr_dat %>% datatable()

avg_overall_dat$type = "Overall"

colnames(avg_overall_dat) <- c("state", "avg_vaccine", "spend_person", "statespending2016", "type"
)

avg_mmr_dat$type ="MMR"

colnames(avg_mmr_dat) <- c("state", "avg_vaccine", "spend_person", "statespending2016", "type"
)

avg_vaccine_dat <- rbind(avg_overall_dat,avg_mmr_dat)

avg_vaccine_dat$type <- factor(avg_vaccine_dat$type,
                               levels = c("Overall","MMR"),
                               labels =c("Overall","MMR") )

```

averaged vaccination rate by state and outcome
```{r}
#| fig-width: 16
#| fig-height: 8
scatter_plot0_overall <- avg_vaccine_dat %>% filter(type=="Overall") %>% 
  ggplot(aes(x = spend_person, y = avg_vaccine)) +
  geom_smooth(method = "lm", se = FALSE,color="gray",linetype="dashed",size=2) +  # One overall trend line
  geom_point(size=5,shape=1) +  
  facet_grid(~type)+
  theme_bw()+
  xlab("Public Spending per Studnet per School Year(in US Dollars)")+
  ylab("Averaged Vaccination Rate in Each State")+
  theme(
    text = element_text(size = 20),          # Overall text size
    axis.text = element_text(size = 18),     # Axis tick labels
    axis.title = element_text(size = 22, face = "bold"), # Axis titles
    legend.text = element_text(size = 18),   # Legend text size
    legend.title = element_text(size = 20, face = "bold"), # Legend title
    strip.text = element_text(size = 20, face = "bold")  # Facet labels
  )

scatter_plot0_mmr <- avg_vaccine_dat %>% filter(type=="MMR") %>% 
  ggplot(aes(x = spend_person, y = avg_vaccine)) +
  geom_smooth(method = "lm", se = FALSE,color="gray",linetype="dashed",size=2) +  # One overall trend line
  geom_point(size=5,shape=1) +  
  facet_grid(~type)+
  theme_bw()+
  xlab("Public Spending per Student per School Year(in US Dollars)")+
  ylab("Averaged Vaccination Rate in Each State")+
  theme(
    text = element_text(size = 20),          # Overall text size
    axis.text = element_text(size = 18),     # Axis tick labels
    axis.title = element_text(size = 22, face = "bold"), # Axis titles
    legend.text = element_text(size = 18),   # Legend text size
    legend.title = element_text(size = 20, face = "bold"), # Legend title
    strip.text = element_text(size = 20, face = "bold")  # Facet labels
  )

scatter_plot0_overall
scatter_plot0_mmr

```


### averaged vaccination rate by county and state

```{r}
avg_overall_dat_county <- full_join(
  vaccine_overall %>% 
  group_by(state,county) %>% 
  summarise(avg_overall = mean(overall)),

vaccine_overall %>% 
  select(state,county,spend_person,statespending2016) %>% unique(),
by= c("state","county")
)

avg_overall_dat_county %>% datatable()

avg_mmr_dat_county <- full_join(
  vaccine_mmr %>% 
  group_by(state,county) %>% 
  summarise(avg_mmr = mean(mmr)),

vaccine_mmr %>% 
  select(state,county,spend_person,statespending2016) %>% unique(),
by= c("state","county")
)
avg_mmr_dat_county %>% datatable()

avg_overall_dat_county$type = "Overall"

colnames(avg_overall_dat_county) <- c("state", "county","avg_vaccine", "spend_person", "statespending2016", "type"
)

avg_mmr_dat_county$type ="MMR"

colnames(avg_mmr_dat_county) <- c("state","county", "avg_vaccine", "spend_person", "statespending2016", "type"
)

avg_vaccine_dat_county <- rbind(avg_overall_dat_county,avg_mmr_dat_county)

avg_vaccine_dat_county$type <- factor(avg_vaccine_dat_county$type,
                               levels = c("Overall","MMR"),
                               labels =c("Overall","MMR") )

```

```{r}
#| fig-width: 16
#| fig-height: 8
scatter_plot3 <- avg_vaccine_dat_county %>% 
  ggplot(aes(x = spend_person, y = avg_vaccine)) +
   geom_smooth(method = "lm", se = FALSE, color = "gray", linetype = "dashed", size = 2) +  # Overall trend line
  geom_point(aes(color = state), size = 5) +  # Points from avg_overall_dat_county
  geom_point(data = avg_vaccine_dat, aes(x = spend_person, y = avg_vaccine), color = "black", size = 5, shape = 1) + 
 facet_wrap(~type) + 
   #geom_line(data = avg_overall_dat, aes(x = spend_person, y = avg_vaccine), color = "black", size = 0.5, shape = 1)
  theme_bw() +
  xlab("Public Spending per Student per School Year (in US Dollars)") +
  ylab("Averaged Vaccination Rate in Each County") +
  theme(
    text = element_text(size = 18),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 20, face = "bold"),
    legend.text = element_text(size = 14,face = "bold"),
    legend.title = element_text(size = 16,face = "bold"),
    strip.text = element_text(size = 18, face = "bold")
  )

scatter_plot3
```


### Maps

#### Overall vaccination map
```{r}
#| output: false
# Load US states map data
us_states <- st_as_sf(states())

```


```{r}
overall_map <- avg_overall_dat %>% 
  mutate(state = tolower(state))

us_states <- us_states %>%
  mutate(NAME = tolower(NAME))

geo_dat <- left_join(us_states, overall_map, by = c("NAME" = "state")) %>% 
  unique()
geo_dat <- st_as_sf(geo_dat)



library(ggplot2)
library(dplyr)

# Create a map with text labels for each state
overall_map_plt <- ggplot(data = geo_dat) +
  geom_sf(aes(fill = avg_vaccine)) +  # Color the states by spending
  # scale_fill_viridis_c() +  # Use the 'viridis' color scale
   scale_fill_gradientn(colors = c("#f4e1d2","#f18973", "#bc5a45", "#9e3d2e"),
                        na.value = "#D3D3D3"  ) + 

  theme_minimal() +  # Use a minimal theme
  labs(
   # title = "Public Education Spending in Each State",
    fill = ""
  ) + 
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) + 
  coord_sf(
    xlim = c(-125, -66),  # Longitude limits (approximate for the U.S.)
    ylim = c(24, 50)      # Latitude limits (approximate for the U.S.)
  ) +
  xlab("")+
  ylab("")+
 # geom_sf_text(aes(label = NAME), size = 1.5, color = "black")+
 theme(
  axis.title = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank(),
  panel.grid = element_blank()
)
```
#### MMR vaccination map
```{r}
mmr_map <- avg_mmr_dat %>%  mutate(state = tolower(state))

us_states <- us_states %>%
  mutate(NAME = tolower(NAME))

geo_dat <- left_join(us_states, mmr_map, by = c("NAME" = "state"))
geo_dat <- st_as_sf(geo_dat)



library(ggplot2)
library(dplyr)

# Create a map with text labels for each state
mmr_map_plt <- ggplot(data = geo_dat) +
  geom_sf(aes(fill = avg_vaccine)) +  # Color the states by spending
  # scale_fill_viridis_c() +  # Use the 'viridis' color scale
   # scale_fill_gradient(low = "#feb236", high = "#ff7b25")+
 scale_fill_gradientn(colors = c("#f4e1d2","#f18973", "#bc5a45", "#9e3d2e"),
                        na.value = "#D3D3D3"  ) +  
  theme_minimal() +  # Use a minimal theme
  labs(
   # title = "Public Education Spending in Each State",
    fill = ""
  ) + 
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) + 
  coord_sf(
    xlim = c(-125, -66),  # Longitude limits (approximate for the U.S.)
    ylim = c(24, 50)      # Latitude limits (approximate for the U.S.)
  ) +
  xlab("")+
  ylab("")+
 # geom_sf_text(aes(label = NAME), size = 1.5, color = "black")+
 theme(
  axis.title = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank(),
  panel.grid = element_blank()
)
  # STUSPS

# map_plts <- ggarrange(overall_map_plt,mmr_map_plt)
# 
# ggsave(map_plts,file=here("figures","map_plts.png"),dpi=300)

ggsave(overall_map_plt,file=here("figures","overall_map_plt.png"),dpi=300)

ggsave(mmr_map_plt,file=here("figures","mmr_map_plt.png"),dpi=300)

overall_map_plt
mmr_map_plt

```


### Fit Multi-level Linear Mixed Effects model: Overall vaccination rates

$$
E(Y_{ijk}|b^{(3)}_{k},b^{(2)}_{jk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i + \beta_{4}\text{State  electoral college}_k + {b}^{(3)}_k\ + {b}^{(2)}_{jk}\ 
$$
where Y is vaccination rate (Overall and MMR) for *i*th school in *j*th county and *k*th state (continuous),
State spending is in US dollars per student of the *k*th state (continuous),
School year of the *i*th school (categorical, with 2017-18 as reference category),
State electoral college is the political party that the *k*th state predominantly voted for in 2017-2018 (categorical, with Swing State (vote fluctuates between parties) as the reference category),
${b}_k$ is the random effect of the *k*th state (level 3),
${b}_jk$ is the random effect for the *j*th county in the *k*th (level 2).


## Fit main model: linear mixed effects model with random intercept for county and state (three-level model)

```{r}

## Fit 3-level LMM model (Table 2)
lmm3 <- lmer(overall_recode ~ spend_person + year_recode + state_political_vote + (1 | state) + (1 | county),
                data = vaccine_overall)
summary(lmm3)

## obtain confidence intervals
confint(lmm3, oldNames = FALSE)

```


## Fit simpler models for comparison 

$$
E(Y_{ijk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i + \beta_{4}\text{State political vote}_k
$$

```{r}

## 1) Fit linear regression model without random effects (base model)
lr <- lm(overall_recode ~ spend_person + year_recode + state_political_vote, data = vaccine_overall)
options(scipen=999)
summary(lr)

```

## 2) Fit linear mixed effects model with random intercept for state (two-level model)

$$
E(Y_{ijk}|b_{k}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i + \beta_{4}\text{State political party}_k + {b}_k\ 
$$

```{r}

## Fit 2-level LMM model
lmm2 <- lmer(overall_recode ~ spend_person + year_recode + state_political_vote + (1 | state), data = vaccine_overall)
summary(lmm2)

## obtain confidence intervals
confint(lmm2, oldNames = FALSE)

```


## Compare the three models using AIC and BIC

```{r}

# extract AIC and BIC from base model
info_crit_base <- data.frame(
            "AIC" = AIC(lr),
            "BIC" = BIC(lr)) %>%
  mutate(Model = "Base model",
         Outcome = "Overall vaccination rate") 

# extract AIC and BIC from two-level model
info_crit2 <- data.frame(
            "AIC" = AIC(lmm2),
            "BIC" = BIC(lmm2)) %>%
  mutate(Model = "Two-level",
         Outcome = "Overall vaccination rate")

# extract AIC and BIC from three-level model
info_crit_table <- data.frame(
  "AIC" = AIC(lmm3),
  "BIC" = BIC(lmm3)) %>%
  mutate(Model = "Three-level",
         Outcome = "Overall vaccination rate") %>%
  # add values from two-level model and base model
  rbind(info_crit2, info_crit_base)
info_crit_table

```

### Check model diagnostics
## Diagnostics for stage-one residuals

```{r}

## unable to extract normalized residuals directly:
## https://stackoverflow.com/questions/73459709/how-does-one-compute-the-normalized-model-residuals-based-via-lme4-mermod-in-r
## https://www.rdocumentation.org/packages/lme4/versions/1.1-36/topics/residuals.merMod

## obtain standardized stage-one residuals (school-level) manually
condmean <- predict(lmm3) # obtain conditional means
y <- vaccine_overall$overall_recode
stg1r <- y - condmean
estsigma2 <- sigma(lmm3)^2
estsigmamat <- estsigma2 * diag(dim(vaccine_overall)[1])
# Cholesky decomposition - this step takes a long time
L <- chol(estsigmamat) 
stg1rstar <- solve(L) %*% stg1r
head(stg1rstar)

## compare raw stage-one residuals and standardized residuals
rtype <- c(rep("Raw", length(stg1r)), rep("Standardized", length(stg1r)))
resid <- c(stg1r, stg1rstar)
stg1resids <- as.data.frame(cbind(rtype, as.numeric(resid)))

# create plot
ggplot(stg1resids, aes(x = resid, fill = rtype)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) +
  labs(y = "Count", x = "Residuals", fill = "Type")

```

## Normal QQ Plot of standardized residuals
```{r}
qqnorm(stg1rstar, pch = 1)
qqline(stg1rstar, col = "steelblue", lwd = 2)

qqnorm(residuals(mod1 , type = "n")) # standardized residual
qqline(residuals(mod1 , type = "n"))

## check normality of model residuals
#qqnorm(residuals(lmm3), ylab="Sample Quantiles for residuals")
#qqline(residuals(lmm3), col="red")

```

## Plot standardized residuals against predicted mean
```{r}
yhat <- predict(lmm3)
plot(y = stg1rstar, x = yhat)

```

## Diagnostics for random effects
```{r}

## check normality of model residuals
qqnorm(residuals(lmm3), ylab="Sample Quantiles for residuals")
qqline(residuals(lmm3), col="red")

```

```{r}
## Normal qqplot of the random effects: county-level random effects
bjkhat <- ranef(lmm3, condVar = TRUE)$county %>% pull()

qqnorm(bjkhat, pch = 1)
qqline(bjkhat, col = "steelblue", lwd = 2)

```

```{r}

## Normal qqplot of the random effects: state-level random effects
bkhat <- ranef(lmm3, condVar = TRUE)$state %>% pull()

qqnorm(bkhat, pch = 1)
qqline(bkhat, col = "steelblue", lwd = 2)

```

```{r}

## Histogram of county random effects residuals
ranef(lmm3)$county %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

```

```{r}

## Histogram of state random effects residuals
ranef(lmm3)$state %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

```

### Intra-class clustering

```{r}
## Table 3

## extract random effects variances
re_variance <- as.data.frame(VarCorr(lmm3))

# create vector for county random effect variance
re_var_county <- re_variance %>% filter(grp=="county") %>% select(vcov) %>% pull()

# create vector for state random effect variance
re_var_state <- re_variance %>% filter(grp=="state") %>% select(vcov) %>% pull()

# create vector for error term variance
re_var_resid <- re_variance %>% filter(grp=="Residual") %>% select(vcov) %>% pull()
  
## correlation between two schools in the same county within the same state
corr_county <- as.data.frame(as.numeric(re_var_county + re_var_state) / (re_var_resid + re_var_state + re_var_county)) %>%
  dplyr::rename(Correlation_coefficient = 1) %>%
  mutate(Type = "Schools in same county")

## correlation between two schools in different counties but within the same state
corr_state <- as.data.frame(as.numeric(re_var_state / (re_var_resid + re_var_state + re_var_county))) %>%
  dplyr::rename(Correlation_coefficient = 1) %>%
  mutate(Type = "Schools in different counties within same state")

## summarise results in table
corr_table <- rbind(corr_county, corr_state) %>%
  select(Type, Correlation_coefficient)
corr_table

```


### Fit Linear Mixed Effects model: MMR vaccination rates

## Fit main model: linear mixed effects model with random intercept for county and state (three-level model)

```{r}

## Fit 3-level LMM model
lmm3mmr <- lmer(mmr_recode ~ spend_person + year_recode + state_political_vote + (1 | state) + (1 | county),
                data = vaccine_mmr)
summary(lmm3mmr)

## obtain confidence intervals
confint(lmm3mmr, oldNames = FALSE)

```

## Fit simpler models for comparison 

```{r}

## 1) Fit linear regression model without random effects (base model)
lr_mmr <- lm(mmr_recode ~ spend_person + year_recode + state_political_vote, data = vaccine_mmr)
summary(lr_mmr)
anova(lr_mmr)

```
## 2) Fit linear mixed effects model with random intercept for state (two-level model)

```{r}

## Fit 2-level LMM model
lmm2mmr <- lmer(mmr_recode ~ spend_person + year_recode + state_political_vote + (1 | state), data = vaccine_mmr)
summary(lmm2mmr)

## obtain confidence intervals
confint(lmm2mmr, oldNames = FALSE)

```

### Compare three-level and two-level models using AIC and BIC

```{r}

## extract AIC and BIC from base model
info_crit_mmr_base <- data.frame(
            "AIC" = AIC(lr_mmr),
            "BIC" = BIC(lr_mmr)) %>%
  mutate(Model = "Base model",
         Outcome = "MMR vaccination rate")

## extract AIC and BIC from two-level model
info_crit_mmr2 <- data.frame(
            "AIC" = AIC(lmm2mmr),
            "BIC" = BIC(lmm2mmr)) %>%
  mutate(Model = "Two-level",
         Outcome = "MMR vaccination rate")

## extract AIC and BIC from three-level model
info_crit_table_mmr <- data.frame(
  "AIC" = AIC(lmm3mmr),
  "BIC" = BIC(lmm3mmr)) %>%
  mutate(Model = "Three-level",
         Outcome = "MMR vaccination rate") %>%
  # add values from two-level model and base model
  rbind(info_crit_mmr2, info_crit_mmr_base)
info_crit_table_mmr

```

## Check model diagnostics
```{r}
## check normality of residuals
qqnorm(residuals(lmm3mmr), ylab="Sample Quantiles for residuals")
qqline(residuals(lmm3mmr), col="red")

```

```{r}
## Normal qqplot of the random effects: county-level random effects
bjkhat <- ranef(lmm3mmr, condVar = TRUE)$county %>% pull()

qqnorm(bjkhat, pch = 1)
qqline(bjkhat, col = "steelblue", lwd = 2)

```


```{r}

## Normal qqplot of the random effects: state-level random effects
bkhat <- ranef(lmm3mmr, condVar = TRUE)$state %>% pull()

qqnorm(bkhat, pch = 1)
qqline(bkhat, col = "steelblue", lwd = 2)

```

```{r}

## Histogram of county random effects residuals
ranef(lmm3mmr)$county %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

```

```{r}

## Histogram of state random effects residuals
ranef(lmm3mmr)$state %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

```

## Intra-class clustering

```{r}

## extract random effects variances
re_variance_mmr <- as.data.frame(VarCorr(lmm3mmr))

# create vector for county random effect variance
re_var_county_mmr <- re_variance_mmr %>% filter(grp=="county") %>% select(vcov) %>% pull()

# create vector for state random effect variance
re_var_state_mmr <- re_variance_mmr %>% filter(grp=="state") %>% select(vcov) %>% pull()

# create vector for error term variance
re_var_resid_mmr <- re_variance_mmr %>% filter(grp=="Residual") %>% select(vcov) %>% pull()
  
## correlation between two schools in the same county within the same state
corr_county_mmr <- as.data.frame(as.numeric(re_var_county_mmr + re_var_state_mmr) / (re_var_resid_mmr + re_var_state_mmr + re_var_county_mmr)) %>%
  dplyr::rename(Correlation_coefficient = 1) %>%
  mutate(Type = "Schools in same county")

## correlation between two schools in different counties but within the same state
corr_state_mmr <- as.data.frame(as.numeric(re_var_state_mmr / (re_var_resid_mmr + re_var_state_mmr + re_var_county_mmr))) %>%
  dplyr::rename(Correlation_coefficient = 1) %>%
  mutate(Type = "Schools in different counties within same state")

corr_table_mmr <- rbind(corr_county_mmr, corr_state_mmr) %>%
  select(Type, Correlation_coefficient)
corr_table_mmr
```

### Supplementary Table 1

```{r}
generate_table <- function(data, filter_conditions) {
  data %>%
    filter(!!!filter_conditions) %>%
    group_by(state) %>%
    
    summarise(
      county_number = n_distinct(county, na.rm = TRUE),
      school_number = length(name),
      
      spending = unique(statespending2016),
      spend_person = unique(spend_person),
      
      school_year_2017_18 = length(year_recode[year_recode == "2017-18"]),
      school_year_2018_19 = length(year_recode[year_recode == "2018-19"]),
      
      mean_overall = round(mean(overall_recode, na.rm = TRUE), 2),
      sd_overall = round(sd(overall_recode, na.rm = TRUE), 2),
      median_overall = round(median(overall_recode, na.rm = TRUE), 2),
      iqr_overall = round(IQR(overall_recode, na.rm = TRUE), 2),
      
      mean_mmr = round(mean(mmr_recode, na.rm = TRUE), 2),
      sd_mmr = round(sd(mmr_recode, na.rm = TRUE), 2),
      median_mmr = round(median(mmr_recode, na.rm = TRUE), 2),
      iqr_mmr = round(IQR(mmr_recode, na.rm = TRUE), 2)
      
    ) %>%
    mutate(across(everything(), ~ replace(., is.na(.), " "))) %>%

    mutate(
      mean_sd_overall = paste0(mean_overall, " (", sd_overall, ")"),
      median_iqr_overall = paste0(median_overall, " (", iqr_overall, ")"),
      mean_sd_mmr = paste0(mean_mmr, " (", sd_mmr, ")"),
      median_iqr_mmr = paste0(median_mmr, " (", iqr_mmr, ")"))
  }

filter_mmr <- list(quote(mmr_recode != -1 | overall_recode != -1), 
                   quote(mmr_recode != " "))

filter_overall <- list(quote(mmr_recode != -1 | overall_recode != -1), 
                       quote(overall_recode != " "))

filter_both <- list(quote(mmr_recode != -1 | overall_recode != -1), 
                    quote(mmr_recode != " "), 
                    quote(overall_recode != " "))

table_mmr <- generate_table(vaccine_dat, filter_mmr)
table_overall <- generate_table(vaccine_dat, filter_overall)

table_combine <- bind_rows(table_mmr, table_overall)

table_combine %>% 
  select(state, county_number, school_number, 
         school_year_2017_18, school_year_2018_19,
         spending, spend_person, 
         mean_sd_overall, median_iqr_overall, 
         mean_sd_mmr, median_iqr_mmr) %>% 
  
  mutate(across(everything(), ~ replace(., . == "  ( )" |
                                           . == "  ( - )" , " "))) %>% 
  
  kable(col.names = c("Outcome variable",
                      "Number of county report",
                      "Number of school report",
                      "2017 - 2018",
                      "2018 - 2019",
                      "Per <br> state", "Per <br> person", 
                      "Mean (SD)", "Median (IQR)",
                      "Mean (SD)", "Median (IQR)"),
        format = "html", escape = FALSE) %>% 
  
  group_rows("MMR (N = 21)", 1, nrow(table_mmr)) %>% 
  group_rows("Overall (N = 17)", 
             nrow(table_mmr) + 1, 
             nrow(table_mmr) + 
             nrow(table_overall)) %>% 
  
add_header_above(c(" " = 1, " " = 2, "School year" = 2,
                     "State spending (in $1,000)" = 2,
                     "Overall Vaccination Rate" = 2, "MMR Vaccination Rate" = 2))
```

### Supplementary Table 2: AIC and BIC to compare models
```{r}

# combine overall vaccination and MMR vaccination rables
info_criterion <- rbind(info_crit_table, info_crit_table_mmr) %>%
  select(Outcome, Model, AIC, BIC) %>%
  # make it pretty
    kable(col.names = c("Outcome",
                        "Model",
                        "AIC",
                      "BIC"),
        format = "html", escape = FALSE) 
info_criterion

```
