---
title: "Fit linear mixed effects model and diagnostics"
format:
  html
---

```{r, message=FALSE}
library(here)
library(tidyverse)
library(tinytex)
library(PupillometryR)
library(DT)
library(lme4)
library(moments) # for skewness() function
#library(MCMCglmm)
#library(brms)
#library(geepack)
library(car)
#library(knitr)
#library(kableExtra)
#library(table1)
library(readr)
library(dplyr)
#library(sf)
library(tigris)
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
theme_set(theme_bw(base_size = 12)) # Use black/white theme and increase font size for all ggplot figures

```

### Data cleaning 

```{r}

## load dataset
vaccine <- read_csv("data/vaccine.csv")

## clean dtaset
vaccine_dat <- vaccine %>% 
  
  # recode vaccination coverage rates with -1 as missing
  mutate(overall_recode = ifelse(overall == -1, NA, overall),
         mmr_recode = ifelse(mmr == -1, NA, mmr)) %>%   
  
  # recode missing or 2017 school year as 2017-2018 (as per dataset description)
  mutate(year_recode = case_when(
    year == "null" ~ "2017-18",
    year == "2017" ~ "2017-18",
    TRUE ~ year)) %>%
  
  # recode school type into public and private schools
  mutate(type_recode = case_when(
    type == "BOCES" | type == "Kindergarten" | type == "Public" ~ "Public",
    type == "Charter" | type == "Nonpublic" | type == "Private" ~ "Private",
    TRUE ~ type)) %>%
  
  # create state spending per pupil variable (in US$1million) --> stops the lmer() warning about different scales
  mutate(spend_person = round(statespending2016/schagepop2016/1000, 2)) %>%
  
 # replicate Nhat's data cleaning for county name
  mutate(county = case_when(
    county == "State Of Illinois" ~ NA_character_,
    county == "BLAINE" ~ "Blaine",
    county == "ADA" ~ "Ada",
    county == "#N/A" ~ NA_character_,
    county == "Oklahoma/Tulsa" ~ "Oklahoma",
    county == "Walwroth" ~ "Walworth",
    county == "adams" ~ "Adams",
    county == "Colorado BOCS" ~ "Denver",
    county == "New Hampshire" ~ NA_character_,
    TRUE ~ county)) %>%
  
  # standardize with uppercase
  mutate(county = toupper(county)) %>%
  mutate(state = toupper(state)) %>%

  # create variable for whether the state has county-level data
  mutate(has_county = if_else(!is.na(county), 1, 0)) %>%
    
  # create unique school id
  arrange(state, county, city, name) %>%
  mutate(school_id = row_number())


## create MMR dataset
vaccine_mmr <- vaccine_dat %>%
  filter(!is.na(mmr_recode)) %>%
  ## log transformation for negative skew
  #mutate(log_mmr = log(mmr_recode)) %>%
  #mutate(sqrt_mmr = sqrt(max(mmr_recode+1) - mmr_recode)) %>%
  mutate(log10_mmr = log10(max(mmr_recode+1) - mmr_recode)) #%>% #this one works best
  #mutate(inverse_mmr = 1/(max(mmr_recode+1) - mmr_recode))

## create overall vaccination dataset
vaccine_overall <- vaccine_dat %>%
  filter(!is.na(overall_recode)) %>%
  ## log transformation for negative skew
  #mutate(log_overall = log(overall_recode)) %>%
  #mutate(sqrt_overall = sqrt(max(overall_recode+1) - overall_recode)) %>%
  mutate(log10_overall = log10(max(overall_recode+1) - overall_recode)) #%>% #this one works best
  #mutate(inverse_overall = 1/(max(overall_recode+1) - overall_recode))

## resource for data transformations for skewness: https://www.datanovia.com/en/lessons/transform-data-to-normal-distribution-in-r/

```

## Examine skewness

```{r, include = FALSE}

## density plots
summary(vaccine_overall$overall_recode)
skewness(vaccine_overall$overall_recode, na.rm = TRUE)
vaccine_overall$overall_recode %>% density() %>% plot()

ggplot(vaccine_overall, aes(x = overall_recode)) +
  geom_density(fill = "skyblue", alpha = 0.7) +
  labs(title = "Kernel Density Plot of Overall Vaccination Rate per School",
       x = "Overall Vaccination Rate",
       y = "Density")

vaccine_overall$log_overall %>% density() %>% plot()
skewness(vaccine_overall$log_overall, na.rm = TRUE)

vaccine_overall$sqrt_overall %>% density() %>% plot()
skewness(vaccine_overall$sqrt_overall, na.rm = TRUE)

vaccine_overall$log10_overall %>% density() %>% plot()
skewness(vaccine_overall$log10_overall, na.rm = TRUE)

vaccine_overall$inverse_overall %>% density() %>% plot()
skewness(vaccine_overall$inverse_overall, na.rm = TRUE)

## Log10 transformation resulted in the least skewness

```


### Fit Linear Mixed Effects model: Overall vaccination rates

## Fit base model (marginal mean response)

$$
E(Y_{ijk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i
$$

```{r, include = FALSE}

## 1a. fit linear regression model
lr <- lm(overall_recode ~ spend_person + year_recode, data = vaccine_overall)
summary(lr)
anova(lr)

# check normality of residuals --> very skewed
#qqnorm(residuals(lr), ylab="Sample Quantiles for residuals")
#qqline(residuals(lr), col="red")


```

```{r}

## 1b. fit linear regression model with log-transformed response
lr_log <- lm(log10_overall ~ spend_person + year_recode, data = vaccine_overall)
summary(lr_log)

# check normality of residuals
#qqnorm(residuals(lr_log), ylab="Sample Quantiles for residuals")
#qqline(residuals(lr_log), col="red")

```


## Fit linear mixed effects model with random intercept for county and state (three-level model)

$$
E(Y_{ijk}|b^{(3)}_{k},b^{(2)}_{jk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i + {b}^{(3)}_k\ + {b}^{(2)}_{jk}\ 
$$
where Y is vaccination rate (Overall and MMR) for *i*th school in *j*th county and *k*th state (continuous),
State spending is in US dollars per student of the *k*th state (continuous),
School year of the *i*th school (categorical, with 2017-18 as reference category),
${b}_k$ is the random effect of the *k*th state (level 3),
${b}_jk$ is the random effect for the *j*th county in the *k*th (level 2).

```{r}

## 2a. fit 3-level LMM model
lmm3 <- lmer(overall_recode ~ spend_person + year_recode + (1 | state) + (1 | county),
                data = vaccine_overall)
#options(scipen=999)
summary(lmm3)

## obtain confidence intervals
#options(scipen=999)
confint(lmm3, oldNames = FALSE)

# check normality of residuals --> very skewed
#qqnorm(residuals(lmm3), ylab="Sample Quantiles for residuals")
#qqline(residuals(lmm3), col="red")

```

```{r}

## 2b. fit 3-level LMM model excluding schools with missing county
vac_overall_county <- vaccine_overall %>%
  filter(has_county==1) %>%
  mutate(spend_person = spend_person / 1000)

lmm3county <- lmer(overall_recode ~ spend_person + year_recode + (1 | state) + (1 | county),
                data = vac_overall_county)
#options(scipen=999)
summary(lmm3county)

## obtain confidence intervals
#options(scipen=999)
confint(lmm3, oldNames = FALSE)


##** results are identical --> can exclude this model **

```

```{r}

## 2c. fit 3-level LMM model with log-transformed response
lmm3log <- lmer(log10_overall ~ spend_person + year_recode + (1 | state) + (1 | county),
                data = vaccine_overall)
#options(scipen=999)
summary(lmm3log)

## obtain confidence intervals
#options(scipen=999)
confint(lmm3log, oldNames = FALSE)

```


## Fit linear mixed effects model with random intercept for state (two-level model)

$$
E(Y_{ijk}|b_{k}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i + {b}_k\ 
$$

```{r}

## 3a. fit 2-level LMM model
lmm2 <- lmer(overall_recode ~ spend_person + year_recode + (1 | state), data = vaccine_overall)
summary(lmm2)

## obtain confidence intervals
#options(scipen=999)
confint(lmm2, oldNames = FALSE)

# check residuals --> very skewed
#qqnorm(residuals(lmm2), ylab="Sample Quantiles for residuals")
#qqline(residuals(lmm2), col="red")

```

```{r}

## 3b. fit 2-level LMM model with log-transformed response
lmm2log <- lmer(log10_overall ~ spend_person + (1 | state), data = vaccine_overall, 
                REML = TRUE, 
                na.action = na.exclude)
summary(lmm2log)

## obtain confidence intervals
#options(scipen=999)
confint(lmm2log, oldNames = FALSE)

```


## Fit linear mixed effects model with random intercept for county and state (three-level model) with covariate state electoral college (Barret's part)

$$ 
E(Y_{ijk}|b^{(3)}_{k},b^{(2)}_{jk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{School year}_i + \beta_{4}\text{State electoral college}_k + {b}^{(3)}_k\ + {b}^{(2)}_{jk}\ +\epsilon_{jk} 
$$
where Y is vaccination rate (Overall and MMR) for *i*th school in *j*th county and *k*th state (rate),
State spending is in US dollars per student of the *k*th state (continuous),
School year of the *i*th school (categorical, with 2017-18 as reference category),
State electoral college of the *k*th state is a categorical variable (Democrat, Republican, swing state),
${b}_k$ is the random effect of the *k*th state (level 3),
${b}_jk$ is the random effect for the *j*th county in the *k*th (level 2).



## Compare three-level and two-level models using AIC and BIC

```{r}

## used this resource: https://stackoverflow.com/questions/70642232/how-to-extract-information-criterions-from-lme4lmer-model-fitted-by-ml-and-c

## extract AIC and BIC from two-level model
info_crit2 <- data.frame(
            "AIC" = AIC(lmm2),
            "BIC" = BIC(lmm2)) %>%
  mutate(Model = "Two-level")

## extract AIC and BIC from three-level model
info_crit_table <- data.frame(
  "AIC" = AIC(lmm3),
  "BIC" = BIC(lmm3)) %>%
  mutate(Model = "Three-level") %>%
  # add values from two-level model
  rbind(info_crit2)
info_crit_table

##** the 3-level model has lowest AIC and BIC **


```

## Check model diagnostics
```{r}

##** Do we need to check stage-one residuals (Cholesky decomposition)? We are not looking at repeated observations among schools. **

# obtain standardized stage-one residuals (school-level)
condmean <- predict(lmm3) # obtain conditional means
y <- vaccine_overall$overall_recode
stg1r <- y - condmean
estsigma2 <- sigma(lmm3)^2
estsigmamat <- estsigma2 * diag(dim(vaccine_overall)[1])
L <- chol(estsigmamat) # this take a long time
stg1rstar <- solve(L) %*% stg1r
head(stg1rstar)

## compare raw stage-one residuals and standardized residuals
rtype <- c(rep("Raw", length(stg1r)), rep("Standardized", length(stg1r)))
resid <- c(stg1r, stg1rstar)
stg1resids <- as.data.frame(cbind(rtype, as.numeric(resid)))

# create plot
ggplot(stg1resids, aes(x = resid, fill = rtype)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) +
  labs(y = "Count", x = "Residuals", fill = "Type")

## QQ Plot of standardized residuals

## Plot standardized residuals against predicted mean

```

```{r}
## check normality of residuals
qqnorm(residuals(lmm3), ylab="Sample Quantiles for residuals")
qqline(residuals(lmm3), col="red")
```

```{r}
## Normal qqplot of the random effects: county-level random effects
bjkhat <- ranef(lmm3, condVar = TRUE)$county %>% pull()

qqnorm(bjkhat, pch = 1)
qqline(bjkhat, col = "steelblue", lwd = 2)

##** skewed on the left **

```


```{r}

## Normal qqplot of the random effects: state-level random effects
bkhat <- ranef(lmm3, condVar = TRUE)$state %>% pull()

qqnorm(bkhat, pch = 1)
qqline(bkhat, col = "steelblue", lwd = 2)

##** looks fairly normal **

```

```{r}

## Histogram of county random effects residuals
ranef(lmm3)$county %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

##** looks fairly normally distributed **

```

```{r}

## Histogram of state random effects residuals
ranef(lmm3)$state %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

##** looks acceptable **

```




## Intra-class clustering

```{r}

## used this resource: https://stackoverflow.com/questions/8526681/extract-random-effect-variances-from-lme4-mer-model-object

## extract random effects variances
re_variance <- as.data.frame(VarCorr(lmm3))

# create vector for county random effect variance
re_var_county <- re_variance %>% filter(grp=="county") %>% select(vcov) %>% pull()

# create vector for state random effect variance
re_var_state <- re_variance %>% filter(grp=="state") %>% select(vcov) %>% pull()

# create vector for error term variance
re_var_resid <- re_variance %>% filter(grp=="Residual") %>% select(vcov) %>% pull()
  
## correlation between two schools in different counties but within the same state
corr_state <- re_var_state / (re_var_resid + re_var_state + re_var_county)
corr_state

## correlation between two schools in the same county within the same state
corr_county_state <- (re_var_county + re_var_state) / (re_var_resid + re_var_state + re_var_county)
corr_county_state

## correlation between two schools in different counties = 0


```



### Fit Linear Mixed Effects model: MMR vaccination rates

## Examine skewness

```{r, include = FALSE}

## density plots
summary(vaccine_mmr$mmr_recode)
skewness(vaccine_mmr$mmr_recode, na.rm = TRUE)
vaccine_mmr$mmr_recode %>% density() %>% plot()

ggplot(vaccine_mmr, aes(x = mmr_recode)) +
  geom_density(fill = "skyblue", alpha = 0.7) +
  labs(title = "Kernel Density Plot of Overall Vaccination Rate per School",
       x = "Overall Vaccination Rate",
       y = "Density")

vaccine_mmr$log_mmr %>% density() %>% plot()
skewness(vaccine_mmr$log_mmr, na.rm = TRUE)

vaccine_mmr$sqrt_mmr %>% density() %>% plot()
skewness(vaccine_mmr$sqrt_mmr, na.rm = TRUE)

vaccine_mmr$log10_mmr %>% density() %>% plot()
skewness(vaccine_mmr$log10_mmr, na.rm = TRUE)

vaccine_mmr$inverse_mmr %>% density() %>% plot()
skewness(vaccine_mmr$inverse_mmr, na.rm = TRUE)

## log10 transformation results in most normal distribution

```


## Fit base model (marginal mean response)

```{r, include = FALSE}

## 4a. fit linear regression model
lr_mmr <- lm(mmr_recode ~ spend_person + year_recode, data = vaccine_mmr)
summary(lr_mmr)
anova(lr_mmr)

# check normality of residuals --> very skewed
#qqnorm(residuals(lr_mmr), ylab="Sample Quantiles for residuals")
#qqline(residuals(lr_mmr), col="red")


```


```{r}

## 4b. fit linear regression model with log-transformed response
lr_mmr_log <- lm(log10_mmr ~ spend_person + year_recode, data = vaccine_mmr)
summary(lr_mmr_log)

# check normality of residuals
#qqnorm(residuals(lr_mmr_log), ylab="Sample Quantiles for residuals")
#qqline(residuals(lr_mmr_log), col="red")

```


## Fit linear mixed effects model with random intercept for county and state (three-level model)

```{r}

## 5a. fit 3-level LMM model
lmm3mmr <- lmer(mmr_recode ~ spend_person + year_recode + (1 | state) + (1 | county),
                data = vaccine_mmr)
#options(scipen=999)
summary(lmm3mmr)

## obtain confidence intervals
#options(scipen=999)
confint(lmm3mmr, oldNames = FALSE)

# check normality of residuals --> very skewed
#qqnorm(residuals(lmm3mmr), ylab="Sample Quantiles for residuals")
#qqline(residuals(lmm3mmr), col="red")

```

```{r}

## 5b. fit 3-level LMM model with log-transformed response
lmm3mmrlog <- lmer(log10_mmr ~ spend_person + year_recode + (1 | state) + (1 | county),
                data = vaccine_mmr)
#options(scipen=999)
summary(lmm3mmrlog)

## obtain confidence intervals
#options(scipen=999)
confint(lmm3mmrlog, oldNames = FALSE)

```

## Fit linear mixed effects model with random intercept for state (two-level model)

```{r}

## 6a. fit 2-level LMM model
lmm2mmr <- lmer(mmr_recode ~ spend_person + year_recode + (1 | state), data = vaccine_mmr)
summary(lmm2mmr)

## obtain confidence intervals
#options(scipen=999)
confint(lmm2mmr, oldNames = FALSE)

# check residuals --> very skewed
#qqnorm(residuals(lmm2mmr), ylab="Sample Quantiles for residuals")
#qqline(residuals(lmm2mmr), col="red")

```

```{r}

## 6b. fit 2-level LMM model with log-transformed response
lmm2mmrlog <- lmer(log10_mmr ~ spend_person + (1 | state), data = vaccine_mmr, 
                REML = TRUE, 
                na.action = na.exclude)
summary(lmm2mmrlog)

## obtain confidence intervals
#options(scipen=999)
confint(lmm2mmrlog, oldNames = FALSE)

```

### Compare three-level and two-level models using AIC and BIC

```{r}

## used this resource: https://stackoverflow.com/questions/70642232/how-to-extract-information-criterions-from-lme4lmer-model-fitted-by-ml-and-c

## extract AIC and BIC from two-level model
info_crit_mmr2 <- data.frame(
            "AIC" = AIC(lmm2mmr),
            "BIC" = BIC(lmm2mmr)) %>%
  mutate(Model = "Two-level")

## extract AIC and BIC from three-level model
info_crit_table_mmr <- data.frame(
  "AIC" = AIC(lmm3mmr),
  "BIC" = BIC(lmm3mmr)) %>%
  mutate(Model = "Three-level") %>%
  # add values from two-level model
  rbind(info_crit_mmr2)
info_crit_table_mmr

##** the 3-level model has lowest AIC and BIC **


```

## Check model diagnostics
```{r}
## check normality of residuals
qqnorm(residuals(lmm3mmr), ylab="Sample Quantiles for residuals")
qqline(residuals(lmm3mmr), col="red")

##** skewed on the left **

```

```{r}
## Normal qqplot of the random effects: county-level random effects
bjkhat <- ranef(lmm3mmr, condVar = TRUE)$county %>% pull()

qqnorm(bjkhat, pch = 1)
qqline(bjkhat, col = "steelblue", lwd = 2)

##** slightly skewed on the left **

```


```{r}

## Normal qqplot of the random effects: state-level random effects
bkhat <- ranef(lmm3mmr, condVar = TRUE)$state %>% pull()

qqnorm(bkhat, pch = 1)
qqline(bkhat, col = "steelblue", lwd = 2)

##** looks fairly normal with single outliers at left and right **

```

```{r}

## Histogram of county random effects residuals
ranef(lmm3mmr)$county %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

##** looks fairly normally distributed around 0 with long left tail **

```

```{r}

## Histogram of state random effects residuals
ranef(lmm3mmr)$state %>%
  pivot_longer(cols = c("(Intercept)"),
               names_to = "ranefs",
               values_to = "estimates") %>%
  ggplot(aes(x = estimates)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  facet_grid(~ ranefs)

##** looks acceptable **

```

## Intra-class clustering

```{r}

## used this resource: https://stackoverflow.com/questions/8526681/extract-random-effect-variances-from-lme4-mer-model-object

## extract random effects variances
re_variance <- as.data.frame(VarCorr(lmm3mmr))

# create vector for county random effect variance
re_var_county <- re_variance %>% filter(grp=="county") %>% select(vcov) %>% pull()

# create vector for state random effect variance
re_var_state <- re_variance %>% filter(grp=="state") %>% select(vcov) %>% pull()

# create vector for error term variance
re_var_resid <- re_variance %>% filter(grp=="Residual") %>% select(vcov) %>% pull()
  
## correlation between two schools in different counties but within the same state
corr_state_mmr <- re_var_state / (re_var_resid + re_var_state + re_var_county)
corr_state_mmr

## correlation between two schools in the same county within the same state
corr_county_state_mmr <- (re_var_county + re_var_state) / (re_var_resid + re_var_state + re_var_county)
corr_county_state_mmr

## correlation between two schools in different counties = 0


```