---
title: "Analysis of Correlated Data (CHL5222H Winter 2025)"
subtitle: "Final Group Project"
author: "Group 5"
toc: true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(tinytex)
library(PupillometryR)
library(DT)
library(lme4)
library(car)
library(knitr)
library(kableExtra)
library(sf)
library(tigris)
library(readr)
library(dplyr)
knitr::opts_knit$set(root.dir = here())
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
theme_set(theme_bw(base_size = 20)) # Use black/white theme and increase font size for all ggplot figures
```

load analytical data:
```{r}
vaccine_overall <- readRDS("~/DLSPH/CHL5222/CHL 5222 Final project/data/vaccine_overall.rds")

vaccine_mmr <- readRDS("~/DLSPH/CHL5222/CHL 5222 Final project/data/vaccine_mmr.rds")
```

before modelling, I looked at the outcome distribution

```{r}
vaccine_overall$overall %>% density() %>% plot()

vaccine_overall$mmr %>% density() %>% plot()

```

apply logit transformation, I have a less skewed distribution
```{r}

p_overall <- (vaccine_overall$overall)*0.01
p_mmr <- (vaccine_overall$mmr)*0.01

p_overall %>% density() %>% plot()

log(  p_overall/ (1-p_overall) ) %>% 
  density() %>% plot()

tmp <- vaccine_mmr %>% 
  mutate(p_mmr = mmr*0.01,
         logit = log(p_mmr/(1-p_mmr)))

# log(  p_mmr/ (1-p_mmr) ) %>% 
#   density() %>% plot()
```

the averaged vaccination rate vs spending
```{r}
avg_overall_dat <- full_join(
  vaccine_overall %>% 
  group_by(state) %>% 
  summarise(avg_overall = mean(overall)),

vaccine_overall %>% 
  select(state,spend_person) %>% unique(),
by= c("state")
)

avg_overall_dat %>% datatable()

avg_mmr_dat <- full_join(
  vaccine_mmr %>% 
  group_by(state) %>% 
  summarise(avg_mmr = mean(mmr)),

vaccine_mmr %>% 
  select(state,spend_person) %>% unique(),
by= c("state")
)
avg_mmr_dat %>% datatable()
```

```{r}
avg_overall_dat %>% 
  ggplot(aes(x = spend_person,y = avg_overall,color = state))+
  geom_point()+
  theme_bw()


avg_mmr_dat %>% 
  ggplot(aes(x = spend_person,y = avg_mmr,color = state))+
  geom_point()+
  theme_bw()

```
    
From the scatter plots, we can see that:      

-  states with similar public spending have different averaged overall vaccination rate, indicating heterogeneity. 
- but I did not observe a trend/relationship between pulic spending and vaccination rate.


### model 1: GLMM without county-level data
the model is specified as:

$$
\text{g}E(Y_{ik}|b^{(3)}_{k}) = \beta_{1} + \beta_{2}\text{State spending}_k + {b}^{(3)}_k\  +\epsilon_{jk}
$$

I normalized the continuous variable

```{r}
colnames(vaccine_overall)

model_overall_dat <- vaccine_overall %>% 
  mutate(p_overall = overall*0.01,
         spend_person_scale = scale(spend_person))

```

```{r}
model_overall_dat %>% 
  ggplot(aes(x = spend_person_scale))+
  geom_density()

model_overall_dat %>% 
  ggplot(aes(x = spend_person))+
  geom_density()

```

```{r}

model1 <- glmer(
  p_overall ~ spend_person_scale+ (1 | state),
   family = gaussian(link = "identity"),
  #  family = binomial("logit"),
  data = model_overall_dat
)

summary(model1)
```

### model 2: GLMM with county-level data

$$
\text{log}E(Y_{ijk}|b^{(3)}_{k},b^{(2)}_{jk}) = \beta_{1} + \beta_{2}\text{State spending}_k + \beta_{3}\text{State electoral college}_k + {b}^{(3)}_k\ + {b}^{(2)}_{jk}\ +\epsilon_{jk}
$$
#### model 2a: with all states

states where there is no county data would not have variation in county-level
```{r}
library(brms)  

model_overall_dat2a <- model_overall_dat %>% 
  filter( county_recode = ifelse(!is.na(county),"county",county ) )


tictoc::tic()
model2b <- brms::brm(p_overall ~ spend_person_scale + (1|state) + (1|county_recode), 
                         data = model_overall_dat2a,
                        # family = "bernoulli", 
                         chains = 4,    # number of chains  
                         iter = 1000, # number of iterations, including burnin 
                         warmup = 200) # burnin  summary(nested.brms)
tictoc::toc()

```
#### model 2b: with states with county-level data
```{r}
model_overall_dat2b <- model_overall_dat %>% 
  filter(!is.na(county))

tictoc::tic()
model2b <- brms::brm(p_overall ~ spend_person_scale + (1|state) + (1|county), 
                         data = model_overall_dat2b,
                        # family = "bernoulli", 
                         chains = 4,    # number of chains  
                         iter = 1000, # number of iterations, including burnin 
                         warmup = 200) # burnin  summary(nested.brms)
tictoc::toc()


```